---
title: "The object manifest"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The object manifest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options("giotto.use_conda" = FALSE, "giotto.no_python_warn" = TRUE)
```

A `giotto` object is large, deeply nested, and mostly opaque from the console.
Two questions come up constantly and neither has a good answer from printed
output:

* **What is in this object?**
* **Did the step I just ran actually change anything?**

`objManifest()` answers the first and `manifestDiff()` answers the second.

```{r, message = FALSE}
library(GiottoClass)

g <- GiottoData::loadGiottoMini("vizgen")
activeSpatUnit(g) <- "aggregate"

objManifest(g)
```

The manifest is what the object **is**. Its companion, `objHistory()`, records
**why** it looks that way. The two are deliberately separate: the history is
provenance, and object state is never reconstructed by replaying it.

## Reading the inventory

The manifest nests exactly as the object nests, so a path through it is a path
through the object — `spat_unit`, then `feat_type`, then name:

```{r}
m <- objManifest(g)
str(m$slots$expression$aggregate$rna$raw)
```

Every accessor used to build a leaf is individually guarded. A field that
cannot be read becomes `NULL` and its path is listed in `warnings`, rather than
erroring the whole manifest — a state report that crashes is worse than no
state report. On a healthy object the list is empty:

```{r}
m$warnings
```

## Did my step do anything?

This is the day-to-day use. Take a manifest before the step, another after,
and ask what moved:

```{r}
before <- objManifest(g, level = "full")

g <- createSpatialNetwork(g, name = "tutorial_net", method = "Delaunay")

manifestDiff(before, objManifest(g, level = "full"))$summary
```

Steps that overwrite something already present are the interesting case.
Recomputing an existing network leaves its name, its position and often its
shape untouched, and only the content changes:

```{r}
before <- objManifest(g, level = "full")

g <- createNearestNetwork(g, name = "sNN.pca", k = 10)

manifestDiff(before, objManifest(g, level = "full"))$summary
```

`manifestDiff()` also returns `changed` (a flag) and `detail` (every affected
leaf, as data). The one-sentence `summary` is what you would hand to a person,
or a model; `detail` is what you would score or assert against.

## `summary` versus `full`

`level = "summary"` describes structure: names, shapes, column names, counts.
`level = "full"` adds a content fingerprint per leaf. That difference matters
exactly when a step rewrites values in place:

```{r}
g2 <- createNearestNetwork(g, name = "sNN.pca", k = 20)

# structure only
manifestDiff(objManifest(g), objManifest(g2))$summary

# structure and content
manifestDiff(objManifest(g, level = "full"),
             objManifest(g2, level = "full"))$summary
```

Fingerprints are cheaper than they sound. They hash a deterministic
fixed-stride sample of the content rather than all of it, so the cost is flat
in object size — about 6 ms for a two-million-point `giottoPoints` and 1 ms for
a matrix with twenty million non-zeros. A full manifest of a real Xenium crop
takes ~44 ms and occupies ~120 KB in memory.

Sampling is stride-based rather than random, so it never touches the RNG and
never disturbs a seeded analysis.

A fingerprint identifies *values*, not their storage. Numbers are compared to
12 significant digits and geometry coordinates as an ordered set of vertices,
so reading back an object you just saved does not report a change merely
because a file format dropped the last bits of a double or renumbered the
vertices of a polygon. The same slack means a difference smaller than about
1e-12 relative is not reported at all.

## The history log

`ghistory_records()` returns the object's history as structured records rather
than printed text:

```{r}
g <- subsetGiotto(g, cell_ids = head(spatIDs(g), 100))

str(tail(ghistory_records(g), 1)[[1]], max.level = 1)
```

`params` holds the arguments the function actually ran with, as deparsed
expressions, so `1:30` survives as `"1:30"`.

Not every change is logged, though. Only a minority of the suite's functions
call `update_giotto_params()`, and a direct slot assignment calls nothing at
all — so an object's contents can move with nothing in its history claiming
responsibility. `recordGiottoStep()` closes that gap for a caller that can see
the change happen:

```{r}
before <- objManifest(g)

cm <- getCellMetadata(g, output = "cellMetaObj")
cm[]$hand_label <- rep(c("tumor", "stroma"), length.out = nrow(cm))
g <- setGiotto(g, cm, verbose = FALSE)

d <- manifestDiff(before, objManifest(g))
d$summary

g <- recordGiottoStep(g, fn = "manual annotation",
                      status = "unattributed", diff = d$detail)

tail(names(objHistory(g)), 1)
```

A failed call is recordable the same way, with `status = "error"`.

## On disk

`saveGiotto()` writes two sidecars beside the object: `manifest.json` (what it
is) and `history.ndjson` (why, one operation per line).

```{r}
td <- file.path(tempdir(), "manifest_vignette")
dir.create(td, recursive = TRUE, showWarnings = FALSE)

saveGiotto(g, dir = td, foldername = "mini", verbose = FALSE)

saved <- file.path(td, "mini")
list.files(saved)
```

The point of the sidecar is that it answers questions about the object without
loading it. Here the manifest is a few tens of kilobytes beside a payload
thousands of times larger, and the ratio only improves as objects grow:

```{r, eval = requireNamespace("jsonlite", quietly = TRUE)}
sc <- jsonlite::fromJSON(file.path(saved, "manifest.json"),
                         simplifyVector = FALSE)

unlist(sc$summary$spat_units)
unlist(sc$summary$n_cells)
names(sc$slots$expression$aggregate$rna)
```

`objManifest_json()` and `objHistory_ndjson()` produce the same two documents
for any object, whether or not you are saving it. The JSON follows a versioned
schema that ships with the package:

```{r}
system.file("schema", "giotto-manifest-0.1.0.json", package = "GiottoClass")
```

Keys are sorted at every level so two manifests of the same state serialize
identically, and `NA`, `NaN` and `Inf` are encoded as strings so that a missing
value stays distinguishable from a not-a-number.

## Beyond GiottoClass

The same two documents appear elsewhere in the suite:

* **GiottoDisk** writes them beside each snapshot in a project directory.
  `snapshotManifest()` and `snapshotHistory()` read them back, so you can see
  what a snapshot holds without loading it.
* **GiottoAgent** attaches `status`, `compact_diff` and `manifest_ref` to every
  execution result, so an assistant driving an analysis reads what a step did
  instead of inferring it from console output.

## What it is good for

* Checking that a step did what you expected — and noticing when it did
  nothing.
* Inspecting a saved object or a snapshot without loading it.
* Regression-testing a pipeline: same inputs should give the same fingerprints.
* Diffing two runs, or the same analysis before and after a parameter change.
* Turning `history.ndjson` into a methods section.
* Giving an assistant, an eval harness or a reviewer machine-readable state
  instead of printed text.

## What it does not do

* **Disk-backed expression gets no fingerprint.** GiottoDisk stores,
  `DelayedArray` and BPCells would have to be materialised to hash, so they
  degrade to no fingerprint rather than pulling the data through memory. A
  change made in place to disk-backed values is therefore invisible to the
  diff.
* **The history is not complete by construction.** It records what functions
  chose to log. Entries written with `toplevel = 1` name
  `update_giotto_params` rather than the calling function — a property of those
  call sites, not of the record.
* **The schema is 0.1.0 and not yet released.** Treat the field set as stable
  in spirit, not frozen.
