---
title: "Giotto spatial geometry classes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Giotto spatial geometry classes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## 1. Overview

*GiottoClass* represents spatial polygons and points respectively as S4 classes
`giottoPolygon` and `giottoPoints`. These objects are built on top of *terra*
`SpatVectors` in order to represent spatial biology. They also respond to many
of terra's generics, with some enhancements. Both objects have a
`spatVector` slot that contains the main information. This common slot is inherited
from the `VIRTUAL` `terraVectData` class, however this naming may be subject to
change as other data representations are added.


The structures of these classes are:

```
terraVectData
\- spatVector

giottoPolygon
\- spatVector           (spatial polygons SpatVector)
\- spatVectorCentroids  (centroids points SpatVector)
\- overlaps             (list of overlapped features info)
\- name                 (object name/spat_unit)
\- unique_ID_cache      (cache of polygon IDs)

giottoPoints
\- spatVector           (spatial points SpatVector)
\- networks             (feature network)
\- feat_type            (object name/feat_type)
\- unique_ID_cache      (cache of feature IDs)
```
Of note are the `unique_ID_cache` slots which are used to get around the memory
usage incurred when frequently accessing the `SpatVectors` attributes. The caches
should be updated after modifications to the object. This is automatically
performed when using `[]` subsetting.


## 2.1 Polygon object creation

`createGiottoPolygon()` accepts a number of inputs, including filepaths to spatial
files such as .wkt, .shp, or .GeoJSON or mask image files of common formats such
as .tif or .png. After reading in these files, they are passed to the respective
methods for `SpatVector` and `SpatRaster` (mask image).

`data.frame-like` inputs are also allowed as long as they are formatted similarly
to *terra's* `matrix` representation of polygon geometries, with at least the
following columns.

- `geom`: integer identifier for each polygon
- `part`: integer identifier for each part of the polygon, in the case of a multi polygon
- `x`: numeric x coordinate of a polygon vertex
- `y`: numeric y coordinate of a polygon vertex
- `hole`: integer identifier for each hole within a polygon
- `poly_ID`: standard Giotto identifier for the polygon. Akin to cell_ID

Additional columns can be included as other attributes, but a set of `poly_ID`s
are required.

*For de-novo generation of simple polygon arrays, see the documentation for
`polyStamp()` and `tessellate()`*

## 2.2 Points object creation

There are relatively fewer input methods for `giottoPoints` since they currently
tend to be provided as flat files as opposed to spatial-specific formats.
`createGiottoPoints()` works with 
with `data.frame-like` inputs with at least columns for x-coordinate, y-coordinate,
and ID information. The first two detected numeric columns are expected to be
x and y respectively. The first character column is assumed to be the feat_IDs.
*terra* `SpatVector` inputs are also accepted. 

A helpful parameter is `split_keyword` which accepts a list of regex keywords
to pull matched features into separate `giottoPoints` objects. This is useful
in situations where multiple feature modalities and/or QC targets are provided
together and would interfere with each other if analyzed together. (see examples
in `createGiottoPoints()` documentation)


## 3. Subsetting

`giottoPoints` and `giottoPolygons` can be subset either by logical vectors
or by ID.
```{r}
library(GiottoClass)
gpoly <- GiottoData::loadSubObjectMini("giottoPolygon")
gpoints <- GiottoData::loadSubObjectMini("giottoPoints")

# full objects
print(gpoly)
print(gpoints)

# subsets
plot(gpoly)
plot(gpoly[c("100210519278873141813371229408401071444", "101161259912191124732236989250178928032", "101488859781016188084173008420811094152")])
print(gpoly[c("100210519278873141813371229408401071444", "101161259912191124732236989250178928032", "101488859781016188084173008420811094152")])
plot(gpoly[c(T,F)])

plot(gpoints)
plot(gpoints[c("Fn1")])
print(gpoints[c("Fn1")])
```

`$` extraction can be used to pull attributes of these objects as vectors
```{r}
head(gpoly$poly_ID)
```

## 4. Conversion to data.table
*GiottoClass* has a set of functions for converting `giottoPoints` and `giottoPolygon`
objects between spatial classes using `as.sf()`, `as.sp()`, `as.stars()`, and
`as.terra()`.
*terra*-based `giottoPoints` and `giottoPolygon` objects can also
be converted into `data.table` for geometry manipulation. This method largely
piggybacks off *terra*'s `as.data.frame()` and adds support for `geom = "XY"`
for polygon geometry.

```{r}
gpoly_dt <- data.table::as.data.table(gpoly, geom = "XY")
gpoints_dt <- data.table::as.data.table(gpoints, geom = "XY")

print(gpoly_dt)
print(gpoints_dt)
```

These tables can then be either re-ingested using `createGiottoPolygon()` and
`createGiottoPoints()` or converted to `SpatVector` using `as.polygons()` or
`as.points()`
```{r}
print(as.polygons(gpoly_dt))

# gpoints currently still requires addition of geom, part, and hole cols
gpoints_dt[, geom := 1:.N]
gpoints_dt[, part := 1]
gpoints_dt[, hole := 0]
print(as.points(gpoints_dt))
```

## 5. Centroids
Centroids information are carried by `giottoPolygon` objects in the `spatVectorCentroids`
slot. The `centroids()` generic from *terra* pulls from this slot if the
information already exists. Otherwise, it calculates and returns a set of
centroids for the polygons as `SpatVector` points. The `append_gpolygon` param
makes it so that the `giottoPolygon` with the centroids info appended is returned
instead.


## 6.1 Overlaps
Overlaps are sets of features overlapped by the polygons. `calculateOverlaps()`
is a generic function that performs this overlapping between polygons and points
and polygons and raster (intensity) data.
```{r}
gpoly@overlaps <- NULL # reset overlaps info
gpoly <- calculateOverlap(gpoly, gpoints, verbose = FALSE)

gimg <- GiottoData::loadSubObjectMini("giottoLargeImage")
gpoly <- calculateOverlap(gpoly, gimg, verbose = FALSE, progress = FALSE)
```


Overlaps are stored as a `list` under the `giottoPolygon` `overlaps` slot,
separated by modalities. This list can be retrieved using `overlaps()`.
In the case of overlapped points geometries, the list items are points `SpatVector`
objects. The `poly_ID` column tracks which polygon is overlapping the point
feature as designated by `feat_ID` and `feat_ID_uniq`. `NA` values mean that
the feature was not overlapped.

For overlaps intensities, the results are stored as `data.tables` in a nested
`list` called `intensity` under the main `overlaps` `list`.
```{r}
print(gpoly@overlaps$rna)
print(gpoly@overlaps$intensity)
```

## 6.2 Overlaps to matrix
Overlap results can be converted into expression matrices using `overlapToMatrix()`
```{r}
# points, rna modality (default)
rna_mat <- overlapToMatrix(gpoly)
print(rna_mat[1:3,1:3])

# intensity, dapi_z0
intens_mat <- overlapToMatrix(gpoly, type = "intensity", feat_info = "dapi_z0")
print(intens_mat[1, 1:3, drop = FALSE])
```

```{r}
sessionInfo()
```

