---
title: "giottoView and giottoSpace: recipes for filtering and reframing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{giottoView and giottoSpace: recipes for filtering and reframing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

## What they are

`giottoView` and `giottoSpace` are **non-destructive recipes** attached to a
`giotto` / `giottoMulti` object that change what subsequent accessors see
without mutating the underlying data.

| | `giottoView` | `giottoSpace` |
|---|---|---|
| Records | filter / crop / sample-selection steps | coordinate-frame transforms (spin, affine, flip, ...) |
| Effect on accessors | narrows which cells survive | reframes spatial geometry of survivors |
| Use case | "show me cluster 3 inside this region" | "rotate sample B by 45° and shift to land next to sample A" |
| Lifetime | recipe object — apply / detach at will | recipe object — apply / detach at will |

They're separate by design: a view is a cell-identity filter (predicates on
metadata or expression, optionally a spatial crop bbox), while a space is a
coordinate transform stack. The same view can be applied in any frame; the
same space can be applied to any cell subset.

## When to use which

* **Filter cells:** `giottoView`. Predicates over metadata or expression,
  optional crop region, optional sample selector.
* **Reposition cells:** `giottoSpace`. Spin / affine / shift / flip / rescale
  composed in order. Sample-keyed so different children of a `giottoMulti`
  can carry different transforms.
* **Both:** apply the view first to pick which cells, then the space to
  decide where to draw them. `materialize()` (or the `view=` / `space=`
  args on getters) accepts both.

## giottoView basics

A view starts empty and accumulates steps via piping:

```{r}
library(GiottoClass)

# Empty view
v <- giottoView()
v
```

### Filtering with `subset()`

The predicate is captured unevaluated and resolved later against
`spatValues()`. Free variables in the predicate (literal scalars, vectors)
are captured eagerly so the view is self-contained:

```{r}
v <- giottoView() |>
    subset(leiden_clus == 1)
v
```

Multiple predicates compose by chaining `subset()` — each step intersects
with the previous survivors:

```{r}
clusters_of_interest <- c(1, 3)

v <- giottoView() |>
    subset(leiden_clus %in% clusters_of_interest) |>
    subset(total_expr > 500)
v
```

### Cropping spatial region

`crop()` records a spatial extent (anything `terra::ext()` can interpret —
SpatExtent, bbox, sf bbox, numeric vector) plus an optional `relation`
controlling intersection semantics:

```{r}
v <- giottoView() |>
    crop(terra::ext(6500, 7000, -4500, -4000))
v
```

`relation` is forwarded to `terra::is.related()` and accepts the full set:
`"intersects"` (default), `"within"`, `"contains"`, `"covers"`,
`"covered_by"`, `"overlaps"`, `"touches"`, `"crosses"`, `"disjoint"`.

`crop()` is not limited to rectangular extents — any geometry that
`terra::is.related()` can evaluate against is accepted. Pass a
`SpatVector` polygon (or an `sf` polygon, WKT, etc.) directly as `y`,
and the same `relation` vocabulary applies:

```{r}
# Build an arbitrary ROI polygon (or pull one from gobject@spatial_info,
# a hand-drawn annotation, an imported GeoJSON, ...)
roi <- terra::vect(
    matrix(c(6500, -4500,
             7100, -4400,
             6900, -4000,
             6500, -4200,
             6500, -4500), ncol = 2, byrow = TRUE),
    type = "polygons"
)

v <- giottoView() |>
    crop(roi, relation = "within")   # cells whose centroid is inside roi
```

Bbox crops take an AABB short-circuit at resolution time; non-rectangular
regions fall through to `terra::is.related()`. Geometry of surviving
subobjects is not modified — only the cell set narrows.

The indirect-usage form (`view = `) works on `crop()` too — same pattern
as `subset()` below — for recording a polygon filter directly onto a
slotted view without first building a standalone recipe:

```{r, eval=FALSE}
g <- crop(g, roi, relation = "within", view = "stripe1")
g <- subset(g, total_expr > 1, view = "stripe1")  # appends to the same view
```

Note that eager `crop()` on a `giotto` / `giottoMulti` is **not**
implemented — it would require coordinated narrowing of every spatial
subobject, images, expression, and metadata. The `view = ` path is
mandatory for now; an eager call errors with a pointer to either the
recipe form or the indirect form.

### Predicate frame vs output frame

A view's `crop()` can carry an additional `space = ` arg that binds the
view's `@space` slot — the **frame the crop region is interpreted in**.
This is a separate concern from any output frame the consumer might
request:

```{r, eval=FALSE}
# Crop region drawn in "rotate" coords; binds view@space.
g <- crop(g, ext(3000, 4000, -3000, -2000), view = "test", space = "rotate")

# Default: predicate uses "rotate" frame to decide cells, output stays
# in NATIVE frame.
getSpatialLocations(g, view = "test")            # native coords, narrowed

# Explicit: predicate uses "rotate" frame, output also in "rotate" frame.
getSpatialLocations(g, view = "test", space = "rotate")  # rotated coords
```

The two arguments are independent:

| | source | what it affects |
|---|---|---|
| **Predicate frame** | `view@space` (set by `crop(..., space = "name")`) | how the recorded crop region is interpreted |
| **Output frame** | explicit `space = ` arg on the getter | what frame the returned coordinates are in |

`view@space` is read at the crop step and used to project the region
into whatever the output frame turns out to be — getters that don't
pass `space = ` keep returning native-frame coords. The behavior is
identical across `getSpatialLocations` / `getCellMetadata` /
`getPolygonInfo` / `getExpression`: same recipe, same cell set.

Note the arg overload: `space = ` on `crop()` binds the view's
predicate frame; `space = ` on transform verbs (`spin()`, `affine()`,
…) records the transform as a step onto the named space recipe. Same
referent (the slotted space `"name"`), different operation per verb.

### Sample selection (giottoMulti only)

When the view will be applied to a `giottoMulti`, `selectSamples()` records
which children participate. The step is resolved first, before any filter
or crop steps:

```{r}
v <- giottoView() |>
    selectSamples("visium_a", "visium_b") |>
    subset(leiden_clus == 1)
v
```

## giottoSpace basics

A space records coordinate transforms keyed by sample. Constructing an
empty space defaults to a single-sample sentinel for use on a plain
`giotto`:

```{r}
s <- giottoSpace()
s
```

Naming the sample binds the space to a specific child for `giottoMulti`
use:

```{r}
s_a <- giottoSpace("visium_a")
s_a
```

### Recording transforms

Standard spatial-transform generics (`spin`, `spatShift`, `affine`,
`flip`, `rescale`, `shear`, `zoom`) get a `giottoSpace` signature that
records the call rather than executing it:

```{r}
s <- giottoSpace("visium_b") |>
    spin(angle = 45) |>
    spatShift(dx = 100, dy = 0)
s
```

Steps are applied in recorded order at resolution time.

### Composing spaces with `+`

Two spaces on the same sample concatenate step lists in order; spaces on
different samples merge into a multi-sample space. This lets you build a
"places visium_a and visium_b next to each other" recipe by composing
per-sample transforms:

```{r}
s_a <- giottoSpace("visium_a")   # leave as-is
s_b <- giottoSpace("visium_b") |>
    spin(angle = 90) |>
    spatShift(dx = 8000, dy = 0)

s_layout <- s_a + s_b
s_layout
```

### Piping after `+` broadcasts — so order matters

A transform piped onto a space is recorded on **every sample the space
currently keys**, not just the most recently added one. That is what makes
`giottoSpace("a") |> spin(30)` work without naming the sample twice, but it
means `+` and a pipe do not commute:

```{r}
# spin lands on BOTH samples: the space keys a and b when spin() is recorded
both <- (giottoSpace("a") + giottoSpace("b")) |> spin(30)
both
```

```{r}
# spin lands on `a` ONLY: b is merged in afterwards, already-recorded steps
# are not retro-applied
a_only <- (giottoSpace("a") |> spin(30)) + giottoSpace("b")
a_only
```

Both are legitimate recipes and neither is a special case — the rule is
simply that recording applies to the current key set. Two habits keep it
predictable:

* build each sample's recipe fully, *then* combine with `+`. This is the
  form used above for `s_layout`, and it is the one to reach for by default.
* if you want a transform to apply to every sample, combine first and pipe
  afterwards — and say so in a comment, because the reader cannot tell a
  deliberate broadcast from an accidental one.

A key with no steps still participates at identity, so `giottoSpace("a")`
is a meaningful thing to add: it puts `a` in the space untransformed rather
than leaving it out.

## Which relations use centroids, and which use geometry

`crop()` narrows by a spatial *relation*, and the relation decides what the
cell is represented by:

| relation | evaluated against | note |
|---|---|---|
| `intersects` (default), `disjoint` | the cell **centroid** | the conventional approximation; a cell whose polygon straddles the region boundary but whose centroid falls outside is not kept |
| `within`, `covered_by`, `contains`, `covers`, `overlaps`, `touches`, `crosses` | the cell **polygon** | area- and boundary-based, so a centroid cannot answer them |

This matters because the two disagree, and the difference is the boundary
band. On the visium mini with a 3000-unit box at the tissue centre,
`intersects` keeps 549 cells and `within` keeps 524 — the 25 cells in
between straddle the edge: centroid inside, polygon not fully contained.

The consequence for practice: a geometry relation needs a polygon source on
the object. If there isn't one, `crop()` says so at resolution time rather
than quietly answering the centroid question instead:

```r
# no @spatial_info on this object
giottoView(g, "w") <- giottoView() |> crop(box, relation = "within")
materialize(g, "w")
#> Error: [crop] relation 'within' must be evaluated against cell geometry,
#> but this object has no polygon source to evaluate it on.
```

The routing is a property of the relation alone, not of where the data is
stored, so the same recipe narrows identically over in-memory objects and
over backed stores.

## Indirect usage: `view = ` / `space = ` on existing generics

The recipe constructors above are explicit and composable, but for the
interactive workflow of "try a filter or transform and save it for
later," the existing eager generics (`subset()`, `spin()`, `spatShift()`,
`flip()`, `rescale()`, `affine()`) accept an optional `view = ` or
`space = ` argument. When supplied, the call records the step onto the
named recipe (creating the slotted view/space if missing) and returns
the modified gobject. When omitted, the call behaves eagerly as before
— no behavior change.

### Recording a view from `subset()`

```{r}
g <- GiottoData::loadGiottoMini("visium", verbose = FALSE)

# Indirect: record the filter under the slot name "cluster1_only"
g <- subset(g, leiden_clus == 1, view = "cluster1_only")

# A second subset() call with the same name appends to that view
g <- subset(g, total_expr > 500, view = "cluster1_only")

# Inspect what was recorded
giottoView(g, "cluster1_only")
```

The eager path is unchanged when `view` is not supplied:

```{r, eval=FALSE}
g_sub <- subset(g, leiden_clus == 1)   # returns a narrowed giotto
```

### Recording a space from a transform

```{r}
# Indirect: record a rotation under the slot name "rotated"
g <- spin(g, angle = 45, space = "rotated")

# Chain another transform onto the same space
g <- spatShift(g, dx = 100, dy = 0, space = "rotated")

giottoSpace(g, "rotated")
```

The eager paths still apply when `space` is not supplied. Eagerly applied
transforms walk subobjects and mutate coordinates in place; recipe-mode
just queues the step.

### Passing a recipe object instead of a name

If you want to build a free-standing recipe (no slotting yet), pass a
`giottoView` / `giottoSpace` object as the `view = ` / `space = ` arg.
The generic appends the step and returns the **recipe** (not the
gobject), so you can keep piping:

```{r}
v0 <- giottoView()
v1 <- subset(g, leiden_clus == 2, view = v0)
v1

# Slot it later when ready
giottoView(g, "cluster2") <- v1
```

### Why the same verbs

The intent is that users keep the verbs they already know. The optional
`view = ` / `space = ` arg lets the same call site choose between "do
it now" (default) and "save the recipe for later." No new
`subsetAsView()` / `recordSpace()` functions to learn.


## Slotting onto a gobject

Views and spaces can live as standalone recipes or be attached to a gobject
by name for repeated use. Slotted views participate in `materialize()` /
getter `view = "name"` lookups, so downstream code can refer to them by
string without rebuilding the recipe each call.

```{r}
g <- GiottoData::loadGiottoMini("visium", verbose = FALSE)

giottoView(g, "cluster1_only") <- giottoView() |>
    subset(leiden_clus == 1)

giottoViews(g)
```

Retrieve by name:

```{r}
giottoView(g, "cluster1_only")
```

Remove with `NULL`:

```{r, eval=FALSE}
giottoView(g, "cluster1_only") <- NULL
```

Spaces follow the same `giottoSpace(g, "name") <- value` /
`giottoSpaces(g)` API:

```{r, eval=FALSE}
giottoSpace(g, "layout") <- s_layout
giottoSpaces(g)
```

## Applying views and spaces

### Via getter `view=` / `space=`

Most data accessors take a `view` argument (and `space`, where coordinates
matter). When supplied, the returned subobject is projected through the
recipe before being handed back:

```{r}
# All cells
all_meta <- getCellMetadata(g, output = "data.table")
nrow(all_meta)

# Narrowed to cluster 1 via view
v <- giottoView() |> subset(leiden_clus == 1)
narrowed <- getCellMetadata(g, output = "data.table", view = v)
nrow(narrowed)
```

The view can be supplied directly (a `giottoView` object) or by the name
of a slotted view (a character string).

### Via `materialize()`

`materialize(g, view, space)` applies the recipe across all relevant
slots at once, returning a narrowed gobject. Useful when several
downstream calls need the same narrowing:

```{r}
g_narrow <- materialize(g, v)
g_narrow

# Subsequent accessors see the narrowed view automatically
length(spatIDs(g_narrow))
```

The `slots = ` argument scopes which slots to narrow — handy when only a
subset of data is needed (e.g. plot functions narrow only the slots they
read).

### Combining view and space

The view runs first to pick cell survivors, then the space transforms the
spatial geometry of those survivors:

```{r, eval=FALSE}
g_out <- materialize(g, view = v, space = s)
```

For plain `getCellMetadata` / `spatValues` calls, `space` is accepted for
API symmetry but is a no-op on tabular subobjects — only `cell_ID`
survival is affected. To get transformed coordinates, route through a
spatial accessor like `getSpatialLocations(g, view = v, space = s)`.

## giottoMulti context

On a `giottoMulti`, views and spaces carry sample selectors:

* `selectSamples("a", "b")` on a view picks which children participate.
* Sample-keyed transforms on a space define per-child coordinate frames
  for cross-sample layouts.

Composition then lets you build a "place these samples together, filtered
to cluster 3" recipe without touching the underlying data:

```{r, eval=FALSE}
v <- giottoView() |>
    selectSamples("visium_a", "visium_b") |>
    subset(leiden_clus == 3)

s <- giottoSpace("visium_a") +
     (giottoSpace("visium_b") |> spatShift(dx = 8000, dy = 0))

mg_out <- materialize(mg, view = v, space = s)
```

The view's surviving cell IDs are computed once across the gmulti; per-child
getter calls downstream reuse that cache.

## Design notes

* **Recipes, not state.** Applying a view doesn't mutate the gobject —
  `materialize()` returns a new gobject (or you pass the view to a getter
  per-call). Detach by simply not passing it.
* **Lazy.** On a disk-backed gobject (via `gsource`), view predicates and
  spatial crops are pushed into the backend's lazy query plan. The actual
  data pull happens only when a downstream call collects.
* **Composable.** `view1 + view2` is reserved for a future release; for now,
  chain `subset()` / `crop()` on a single view. Spaces compose freely via
  `+` (same-sample concat / cross-sample merge).
* **Sample-keyed transforms.** On a `giottoMulti`, every space step is
  scoped to one or more named samples. Different children carry different
  recipes simultaneously in a single space object.

## See also

* [giottoMulti vignette](giottoMulti.html) — multi-sample container that
  view and space integrate with most deeply.
* `?materialize` — the full resolver entry point.
* `?giottoView` — slotted view accessors.
* `?giottoSpace` — slotted space accessors.
