---
title: "giottoView and giottoSpace: recipes for filtering and reframing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{giottoView and giottoSpace: recipes for filtering and reframing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

## What they are

`giottoView` and `giottoSpace` are **non-destructive recipes** attached to a
`giotto` / `giottoMulti` object that change what subsequent accessors see
without mutating the underlying data.

| | `giottoView` | `giottoSpace` |
|---|---|---|
| Records | filter / crop / sample-selection steps | coordinate-frame transforms (spin, affine, flip, ...) |
| Effect on accessors | narrows which cells survive | reframes spatial geometry of survivors |
| Use case | "show me cluster 3 inside this region" | "rotate sample B by 45° and shift to land next to sample A" |
| Built by | `subset()` / `crop()` / `selectSamples()` with `view = "name"` | a transform verb with `space = "name"` |
| Stored as | a plain list under `gobject@view$name` | a plain list under `gobject@spaces$name` |

Both are **plain lists**, not S4 objects, and both are created by *recording*
onto a name: there is no constructor and no standalone builder. The first call
naming a view or space creates it; later calls with the same name append.

They're separate by design: a view is a cell-identity filter (predicates on
metadata or expression, optionally a spatial crop bbox), while a space is a
coordinate transform stack. The same view can be applied in any frame; the
same space can be applied to any cell subset.

## When to use which

* **Filter cells:** `giottoView`. Predicates over metadata or expression,
  optional crop region, optional sample selector.
* **Reposition cells:** `giottoSpace`. Spin / affine / shift / flip / rescale
  composed in order. Sample-keyed so different children of a `giottoMulti`
  can carry different transforms.
* **Both:** apply the view first to pick which cells, then the space to
  decide where to draw them. `materialize()` (or the `view=` / `space=`
  args on getters) accepts both.

## Recording a view

`subset()` gains a `view = ` argument. With it, the call records a filter
step under that name instead of narrowing the object, and returns the
gobject:

```{r}
library(GiottoClass)

g <- GiottoData::loadGiottoMini("visium", verbose = FALSE)

g <- subset(g, leiden_clus == 1, view = "cluster1")
giottoViews(g)
giottoView(g, "cluster1")
```

The predicate is captured unevaluated and resolved later against
`spatValues()`. Free variables in it (literal scalars, vectors) are
substituted in at record time, so the recipe is self-contained and does not
change if the binding does:

```{r}
clusters_of_interest <- c(1, 3)

g <- subset(g, leiden_clus %in% clusters_of_interest, view = "multi")
giottoView(g, "multi")$steps[[1]]$predicate
```

Naming the same view again appends — each step intersects with the previous
survivors:

```{r}
g <- subset(g, total_expr > 500, view = "multi")
length(giottoView(g, "multi")$steps)
```

`negate = TRUE` is folded into the recorded predicate rather than kept as a
separate field, so the step always reads as the question it will ask:

```{r}
g <- subset(g, leiden_clus == 1, negate = TRUE, view = "not_c1")
giottoView(g, "not_c1")$steps[[1]]$predicate
```

### Cropping spatial region

`crop()` records a spatial extent (anything `terra::ext()` can interpret —
SpatExtent, bbox, sf bbox, numeric vector) plus an optional `relation` and
`geom`:

```{r}
g <- crop(g, terra::ext(6500, 7000, -4500, -4000), view = "box")
giottoView(g, "box")$steps[[1]]$relation
```

`relation` is the spatial predicate; `geom` says what represents a cell when
that predicate runs. Both are recorded on the step — see
[What a crop predicate is evaluated against](#what-a-crop-predicate-is-evaluated-against)
for which relations are available and how the two interact.

`crop()` is not limited to rectangular extents — any geometry that
`terra::is.related()` can evaluate against is accepted. Pass a
`SpatVector` polygon (or an `sf` polygon, WKT, etc.) directly as `y`,
and the same `relation` and `geom` arguments apply:

```{r}
# Build an arbitrary ROI polygon (or pull one from gobject@spatial_info,
# a hand-drawn annotation, an imported GeoJSON, ...)
roi <- terra::vect(
    matrix(c(6500, -4500,
             7100, -4400,
             6900, -4000,
             6500, -4200,
             6500, -4500), ncol = 2, byrow = TRUE),
    type = "polygons"
)

# cells whose centroid is inside roi
g <- crop(g, roi, relation = "within", view = "roi")
```

Bbox crops take an AABB short-circuit at resolution time; non-rectangular
regions fall through to a `spatRelate()` call. Geometry of surviving
subobjects is not modified — only the cell set narrows.

Views mix step kinds freely, since every verb records onto the same name:

```{r, eval=FALSE}
g <- crop(g, roi, relation = "within", geom = "poly", view = "stripe1")
g <- subset(g, total_expr > 1, view = "stripe1")  # appends to the same view
```

Note that eager `crop()` on a `giotto` / `giottoMulti` is **not**
implemented — it would require coordinated narrowing of every spatial
subobject, images, expression, and metadata. The `view = ` path is
mandatory for now; an eager call errors with a pointer to it.

### Predicate frame vs output frame

A view's `crop()` can carry an additional `space = ` arg that binds the
view's `space` field — the **frame the crop region is interpreted in**.
This is a separate concern from any output frame the consumer might
request:

```{r, eval=FALSE}
# Crop region drawn in "rotate" coords; binds the view's `space`.
g <- crop(g, ext(3000, 4000, -3000, -2000), view = "test", space = "rotate")

# Default: predicate uses "rotate" frame to decide cells, output stays
# in NATIVE frame.
getSpatialLocations(g, view = "test")            # native coords, narrowed

# Explicit: predicate uses "rotate" frame, output also in "rotate" frame.
getSpatialLocations(g, view = "test", space = "rotate")  # rotated coords
```

The two arguments are independent:

| | source | what it affects |
|---|---|---|
| **Predicate frame** | the view's `space` field (set by `crop(..., space = "name")`) | how the recorded crop region is interpreted |
| **Output frame** | explicit `space = ` arg on the getter | what frame the returned coordinates are in |

The view's `space` is read at the crop step and used to project the region
into whatever the output frame turns out to be — getters that don't
pass `space = ` keep returning native-frame coords. The behavior is
identical across `getSpatialLocations` / `getCellMetadata` /
`getPolygonInfo` / `getExpression`: same recipe, same cell set.

Note the arg overload: `space = ` on `crop()` binds the view's
predicate frame; `space = ` on transform verbs (`spin()`, `affine()`,
…) records the transform as a step onto the named space recipe. Same
referent (the slotted space `"name"`), different operation per verb.

### Sample selection (giottoMulti only)

When the view will be applied to a `giottoMulti`, `selectSamples()` records
which children participate. The step is resolved first, before any filter
or crop steps:

```{r, eval=FALSE}
mg <- selectSamples(mg, "visium_a", "visium_b", view = "pair")
mg <- subset(mg, leiden_clus == 1, view = "pair")
```

## Recording a space

The spatial-transform generics (`spin`, `spatShift`, `affine`, `flip`,
`rescale`) take `space = "name"` and record the call instead of executing
it. On a plain `giotto` there is one sample, so the recipe keys a single
`:default:` sentinel:

```{r}
g <- spin(g, angle = 45, space = "rotated")
g <- spatShift(g, dx = 100, dy = 0, space = "rotated")
str(giottoSpace(g, "rotated")$samples, max.level = 2)
```

Steps apply in recorded order at resolution time. Calling a transform
without `space = ` still does what it always did — transform the object
eagerly.

### Scoping to samples on a `giottoMulti`

On a `giottoMulti`, `samples = ` says which children a transform applies
to. This is how a cross-sample layout is built — one call per child,
against the same space name:

```{r, eval=FALSE}
# leave visium_a where it is; move visium_b next to it
mg <- spin(mg, angle = 90, space = "layout", samples = "visium_b")
mg <- spatShift(mg, dx = 8000, dy = 0, space = "layout",
    samples = "visium_b")
```

`samples = ` accepts several children at once, and omitting it means
"every sample this space already keys" — the whole-layout case:

```{r, eval=FALSE}
mg <- spin(mg, 30, space = "layout", samples = c("visium_a", "visium_b"))
mg <- spatShift(mg, dx = 10, space = "layout")   # applies to both
```

A mistyped child name is rejected at record time rather than quietly
creating a chain nothing resolves against.

`samples = ` exists only on `giottoMulti` methods. A plain `giotto` has one
sample, so there is nothing to scope to, and a gmulti transform *requires*
`space = ` — eager per-child transforms are not implemented, and the
cross-sample layout that motivates them is exactly what a recorded space
expresses instead.

Earlier versions built spaces standalone and merged them with `+`. That is
gone: `+` merged key sets, so a transform recorded afterwards landed on
whatever had been merged so far, and the same expression meant different
things depending on build order. `samples = ` states the scope at each call
instead.

## What a crop predicate is evaluated against

A crop narrows the **cell set**, so each cell must be reduced to a geometry
before the predicate can run. `geom` declares which one:

* `geom = "centroid"` (default) — the cell's `spatial_locs` row. Cheap, and
  the conventional choice in spatial omics; `combineCellData()` documents the
  same. Approximate: a cell whose polygon straddles the region boundary with
  its centroid outside is dropped.
* `geom = "poly"` — the cell's actual polygon. Exact, and needs a polygon
  source on the object.

Declaring it rather than inferring it from the relation matters because the
two genuinely answer different questions, and a saved recipe should record
which one it asks. Same relation, both representations, on the visium mini
with a 3000-unit box at the tissue centre (624 cells total):

| relation | `geom = "centroid"` | `geom = "poly"` |
|---|---|---|
| `intersects` | 549 | 549 |
| `within` | 549 | **524** |

The `within` row is the interesting one: 25 spots have their centroid inside
the box but are not *wholly* inside it. Whether you want those is a question
about your analysis, not something a relation name can settle. (`within` on a
centroid coincides with `intersects` on a centroid here, because for a point
the two differ only when it lands exactly on the boundary.)

Not every relation works on a centroid. `contains`, `covers`, `overlaps` and
`crosses` are always `FALSE` against a point — a point cannot contain a
polygon — so requesting one promotes `geom` to `"poly"` and says so:

```{r}
g <- crop(g, roi, relation = "contains", view = "contains_demo")
giottoView(g, "contains_demo")$steps[[1]]$geom
```

`intersects`, `disjoint`, `within` and `touches` are all well defined on a
centroid and are left alone. (`covered_by` reads like a predicate but terra
does not accept it, so it is rejected.)

Declaring `geom = "poly"` on an object with no polygon source is an error
rather than a silent fallback to the centroid answer:

```r
# no @spatial_info on this object
g_nopoly <- crop(g_nopoly, box, geom = "poly", view = "p")
materialize(g_nopoly, "p")
#> Error: [crop] geom = "poly" was requested (relation 'intersects'), but
#> this object has no polygon source to evaluate it on.
#> Either add polygons (`setPolygonInfo()`) or use geom = "centroid".
```

Because the choice lives on the step instead of being re-derived per
substrate, one recipe narrows identically across every cell-keyed slot and
across in-memory objects and backed stores.

## `view = ` and `space = ` take a name, not a recipe

Both are `character(1)` — the key the recipe lives under. Passing a recipe
inline was considered and deliberately rejected (see
`vignettes/articles/DESIGN_gmulti_federation.md`): recipes are curated,
named artifacts, and accepting one inline would make the same call site
sometimes return a gobject and sometimes return a recipe.

There is no separate constructor to reach for, because recording *is* the
construction path — the first call naming a view or space creates it. So
the accessors exist to inspect, copy, and remove, not to build:

```{r}
giottoView(g, "cluster1")          # read one back
giottoViews(g)                     # list the names
```

The setter's remaining job is moving a recipe between objects, or dropping
one. It validates what it is handed, since the recipe is a plain list that
nothing else type-checks:

```{r, eval=FALSE}
giottoView(g2, "cluster1") <- giottoView(g, "cluster1")   # copy across
giottoView(g, "cluster1") <- NULL                          # remove
```

### Why the same verbs
### Why the same verbs

The intent is that users keep the verbs they already know. The optional
`view = ` / `space = ` arg lets the same call site choose between "do
it now" (default) and "save the recipe for later." No new
`subsetAsView()` / `recordSpace()` functions to learn.


## Where recipes live

A recipe lives on the gobject from the moment it is recorded — under
`@view$name` or `@spaces$name` — and travels with it through
`saveGiotto()` / `loadGiotto()`. Downstream code refers to it by string, so
nothing needs rebuilding per call:

```{r}
g <- GiottoData::loadGiottoMini("visium", verbose = FALSE)
g <- subset(g, leiden_clus == 1, view = "cluster1_only")

giottoViews(g)
giottoView(g, "cluster1_only")
```

Several recipes coexist on one object, which is the point: an object can
carry competing narrowings and frames at once, and each accessor call picks
one by name.

## Applying views and spaces

### Via getter `view=` / `space=`

Most data accessors take a `view` argument (and `space`, where coordinates
matter). When supplied, the returned subobject is projected through the
recipe before being handed back:

```{r}
# All cells
all_meta <- getCellMetadata(g, output = "data.table")
nrow(all_meta)

# Narrowed to cluster 1 via the view slotted above
narrowed <- getCellMetadata(g, output = "data.table",
    view = "cluster1_only")
nrow(narrowed)
```

The view is named, not passed inline — see
[`view = ` and `space = ` take a name, not a recipe](#view---and-space---take-a-name-not-a-recipe)
above.

### Via `materialize()`

`materialize(g, view, space)` applies the recipe across all relevant
slots at once, returning a narrowed gobject. Useful when several
downstream calls need the same narrowing:

```{r}
g_narrow <- materialize(g, "cluster1_only")
g_narrow

# Subsequent accessors see the narrowed view automatically
length(spatIDs(g_narrow))
```

The `slots = ` argument scopes which slots to narrow — handy when only a
subset of data is needed (e.g. plot functions narrow only the slots they
read).

### Combining view and space

The view runs first to pick cell survivors, then the space transforms the
spatial geometry of those survivors:

```{r, eval=FALSE}
g <- spatShift(g, dx = 100, dy = 0, space = "layout")
g_out <- materialize(g, view = "cluster1_only", space = "layout")
```

For plain `getCellMetadata` / `spatValues` calls, `space` is accepted for
API symmetry but is a no-op on tabular subobjects — only `cell_ID`
survival is affected. To get transformed coordinates, route through a
spatial accessor like
`getSpatialLocations(g, view = "cluster1_only", space = "layout")`.

## giottoMulti context

On a `giottoMulti`, views and spaces carry sample scope:

* `selectSamples(mg, "a", "b", view = )` picks which children participate.
* `samples = ` on a transform defines per-child coordinate frames for
  cross-sample layouts.

Together they build a "place these samples together, filtered to cluster 3"
recipe without touching the underlying data:

```{r, eval=FALSE}
mg <- selectSamples(mg, "visium_a", "visium_b", view = "cluster3")
mg <- subset(mg, leiden_clus == 3, view = "cluster3")

# visium_a stays put; visium_b moves next to it
mg <- spatShift(mg, dx = 8000, dy = 0, space = "side_by_side",
    samples = "visium_b")

mg_out <- materialize(mg, view = "cluster3", space = "side_by_side")
```

The view's surviving cell IDs are computed once across the gmulti; per-child
getter calls downstream reuse that cache.

## Design notes

* **Recipes, not state.** *Resolving* a view doesn't mutate the gobject —
  `materialize()` returns a new gobject, or you pass the name to a getter
  per-call. Recording a step is the one write, and it touches only the
  recipe, never the data.
* **Plain lists.** A view is `list(steps = , space = , misc = )`; a space is
  `list(samples = , misc = )`. Neither is S4. That is what lets a recipe
  survive `saveRDS()` and reach a parallel worker: the predicate is stored
  deparsed with its environment's values substituted in, and a crop region
  is stored as WKT rather than a terra pointer.
* **Lazy.** On a disk-backed gobject (via `gsource`), view predicates and
  spatial crops are pushed into the backend's lazy query plan. The actual
  data pull happens only when a downstream call collects.
* **Composable by name.** Chain `subset()` / `crop()` calls against the same
  `view = ` to compose; the steps intersect. Cross-*view* composition
  (combining two named views) is reserved for a future release.
* **Sample-scoped transforms.** On a `giottoMulti`, `samples = ` scopes each
  space step to named children, so one space carries different frames for
  different children.

## See also

* [giottoMulti vignette](giottoMulti.html) — multi-sample container that
  view and space integrate with most deeply.
* `?materialize` — the full resolver entry point.
* `?giottoView` — slotted view accessors.
* `?giottoSpace` — slotted space accessors.
